---
title: "Week 10: Multilevel regression and post-stratification"
date: today
date-format: "DD/MM/YY"
format: pdf
number-sections: true
execute: 
  echo: true
  warning: false
  message: false
---

# Overview

In this lab we will look at an example of doing multilevel regression with the fertility intentions data. We will also go through how to do post-stratification and MRP. 


# Packages and data

Read in the data from [NSFG](https://www.cdc.gov/nchs/nsfg/index.htm):

```{r}
library(tidyverse)
library(lme4)
library(rstan)

d <- read_csv("../data/intentions.csv")
d <- d |> 
  mutate(educ_cat = ifelse(edu == "Bachelor"|edu == "Post-Grad", "Bachelor+", edu)) |> 
  mutate(educ_cat = fct_relevel(educ_cat, "<HS", "HS", "Some College", "Bachelor+")) 
```

# Descriptives

Let's do some brief descriptive plots. Here's the observed proportion of women wanting more kids by edication and age:

```{r}
d |>  
  group_by(age_gp, educ_cat, wants_more_kids) |> 
  tally() |> 
  group_by(age_gp, educ_cat) |> 
  summarize(prop_wants_more = n[wants_more_kids==1]/sum(n), n = sum(n)) |> 
  filter(age_gp!="15-19") |> 
  ggplot(aes(age_gp, prop_wants_more, fill = educ_cat)) +
  labs(title = "Observed proportion of women wanting more kids", x = "age", y = "proportion")+
  theme_bw()+
  scale_fill_brewer(palette = "Set1", name = "education")+
  geom_bar(stat = "identity", position = "dodge")
```

Here's a tile plot of the sample sizes:

```{r}
d |> group_by(age_gp, marital_formal, educ_cat) |> 
  summarize(n = n()) |> 
  filter(age_gp!="15-19") |> 
  ggplot(aes(x=age_gp, y=marital_formal, fill=n)) + 
  geom_tile() + 
  scale_fill_gradient(low="white", high="grey") +
  geom_text(aes(label=n), size=3) +
  xlab("Age Group") + ylab("Marital Status") + theme_minimal()+facet_wrap(~educ_cat)
```

## Question

Do another interesting descriptive chart of the data. 

```{r}
d |> 
  filter(intention < 10) |> 
  mutate(how_many_more = intention - parity) |> 
  ggplot(aes(parity, how_many_more)) + 
  geom_point()+
  facet_wrap(~educ_cat)
```


# Multilevel model

Let's fit the same multilevel model as seen in the lecture using the `lme4` package

```{r}
resp4 <- d |> filter(age_gp!="15-19")  |> drop_na() |> mutate(age_gp = as_factor(age_gp)) |> 
  mutate(age_gp = fct_relevel(age_gp, c("20-24", "25-29", "30-34", "35-39", "40-44")))

mod <- glmer(wants_more_kids ~ marital_formal + (1|age_gp) + (1|educ_cat), data = resp4, family = "binomial")
summary(mod)

mod_parity <- glmer(wants_more_kids ~ marital_formal + (1|age_gp) + (parity|educ_cat), data = resp4, family = "binomial")
summary(mod_parity)
ranef(mod_parity)
```

Note! This can also be done in Stan, in a Bayesian framework. This is useful if you want to e.g. play around with the structure of the hierarchical model, which is much easier to control in this framework. 

```{r}
stan_data1 <- list(N = nrow(resp4),
                  A = nlevels(resp4$age_gp),
                  E = nlevels(resp4$educ_cat), 
                  age = as.numeric(resp4$age_gp),
                  edu = as.numeric(resp4$educ_cat), 
                  x1 = as.numeric(resp4$marital_formal == 'Married'),
                  x2 = as.numeric(resp4$marital_formal == 'Never Married'), 
                  y = resp4$wants_more_kids)

mod1 <- stan(data = stan_data1, 
             file = "../code/models/intentions.stan",
             iter = 500,
             seed = 243)
summary(mod1)$summary[c("beta0", "beta1", "beta2", "sigma_a", "sigma_e"),
                      c("mean", "se_mean", "n_eff", "Rhat")]
```

To pull out the estimates from `lme4`:

```{r}
mod
coef(mod)
fixef(mod)
ranef(mod)
```

## Question

Compare the multilevel model with just a standard logistic regression. Which model has a higher AIC?

```{r}
summary(glm(wants_more_kids ~ marital_formal + age_gp + educ_cat, data = resp4, family = "binomial"))
```


# Post stratification

We are interested in obtaining estimates for the proportion of women who want more kids by age. 

## Raw proportions by age

Before we do any post-stratification, we can just calculate the raw survey proportions by age:

```{r}
raw_age <- resp4 |> 
  mutate(edu = ifelse(edu == "Bachelor"|edu=="Post-Grad", "Bachelor+", as.character(edu))) |> 
  group_by(age_gp) |> 
  summarize(proportion = mean(wants_more_kids, na.rm = TRUE)) |> 
  mutate(estimate = "raw")
```

## Post-stratification dataset

We are going to use data from the American Community Survey (ACS) as the post-stratification dataset. Note that I got these data from IPUMS USA. Load in ACS data:

```{r}
cell_counts <- read_csv("../data/acs_counts.csv")
```

## Question

Compare proportions by age/education/marital status group in ACS versus survey. 

```{r}
cell_counts <- cell_counts |> 
  mutate(prop = n / sum(n))

sum(cell_counts$prop)

survey_counts <- resp4 |> 
  ungroup() |> 
  group_by(age_gp, educ_cat, marital_formal) |> 
  tally() |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

sum(survey_counts$prop)

cell_counts |> 
  rename(prop_acs = prop,
         n_acs = n) |> 
  left_join(survey_counts |> rename(marital = marital_formal, 
                                    edu = educ_cat)) |> 
  select(age_gp, edu, marital, prop_acs, prop) |> 
  pivot_longer(prop_acs:prop) |> 
  ggplot(aes(age_gp, value, fill = name)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  facet_grid(edu~marital)
```


## Standard post-stratification

Calculate the proportion by group (age, education, marital status) and then re-weight based on the ACS data:

```{r}
ps_age <- resp4 |> 
  mutate(edu = ifelse(edu == "Bachelor"|edu=="Post-Grad", "Bachelor+", as.character(edu))) |> 
  group_by(age_gp, edu, marital) |> 
  summarize(prop = mean(wants_more_kids, na.rm = TRUE)) |> 
  filter(marital !="Cohabitating") |> 
  left_join(cell_counts |> select(-prop)) |> 
  mutate(prod = prop*n) |> 
  group_by(age_gp) |> 
  summarize(proportion = sum(prod)/sum(n)) |> 
  mutate(estimate = "poststrat")
```



## MRP

Get predicted proportions for each group:

```{r}
age_groups <- rownames(coef(mod)$age_gp)
educ_groups <- rownames(coef(mod)$educ_cat)
marital_groups <- c("Formerly Married","Married", "Never Married")

pred_data <- expand_grid(age_gp = age_groups, educ_cat = educ_groups, marital_formal = marital_groups)

pred_mrp <- plogis(predict(mod, pred_data))

pred_prop <- cbind(pred_data, pred_mrp)
```

Now post-stratify by age:

```{r}
mrp_age <- pred_prop |> 
  rename(edu = educ_cat, marital = marital_formal) |> 
  left_join(cell_counts |> select(-prop)) |> 
  mutate(prod = pred_mrp*n) |> 
  group_by(age_gp) |> 
  summarize(proportion = sum(prod)/sum(n)) |> 
  mutate(estimate = "mrp")
```

Compare on a plot:

```{r}
ps_age |> 
  bind_rows(mrp_age) |> 
  bind_rows(raw_age) |> 
  ggplot(aes(age_gp, proportion, fill = estimate))+
  geom_bar(stat = "identity", position = "dodge")
```

## Question

Do the same thing but compare across education groups.
