---
title: "Week 3: Mortality II"
subtitle: "SOC6708 ADA"
author: "Monica Alexander"
format: pdf
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(here)
library(readxl)
library(janitor)
```

# Decomposition

Let's read in WPP data from the first week week and calculate the age-specific mortality rates:

```{r}
d_male <- read_xlsx(here("data/WPP2024_POP_F01_2_POPULATION_SINGLE_AGE_MALE.xlsx"), skip = 16)
d_male$sex <- "Male"
d_male <- d_male |> drop_na(Year)
d_female <- read_xlsx(here("data/WPP2024_POP_F01_3_POPULATION_SINGLE_AGE_FEMALE.xlsx"), skip = 16)
d_female$sex <- "Female"
d_female <- d_female |> drop_na(Year)

d <- rbind(d_male, d_female)
rm(d_male, d_female)

d <- d  |> 
  clean_names() |> 
  select(region_subregion_country_or_area, iso3_alpha_code, year, x0:sex) |> 
  rename(region = region_subregion_country_or_area) |> 
  mutate(across(x0:x100, as.numeric))

d_male <- read_xlsx(here("data/WPP2024_MORT_F01_2_DEATHS_SINGLE_AGE_MALE.xlsx"), skip = 16)
d_male$sex <- "Male"
d_male <- d_male |> drop_na(Year)
d_female <- read_xlsx(here("data/WPP2024_MORT_F01_3_DEATHS_SINGLE_AGE_FEMALE.xlsx"), skip = 16)
d_female$sex <- "Female"
d_female <- d_female |> drop_na(Year)

dm <- rbind(d_male, d_female)
rm(d_male, d_female)

dm <- dm  |> 
  clean_names() |> 
  select(region_subregion_country_or_area, iso3_alpha_code, year, x0:sex) |> 
  rename(region = region_subregion_country_or_area) |> 
  mutate(across(x0:x100, as.numeric))

d_long <- d  |>  
  pivot_longer(x0:x100, names_to = "age", values_to = "pop") |> 
  mutate(age = as.numeric(str_remove(age, "x")))

dm_long <- dm  |>  
  pivot_longer(x0:x100, names_to = "age", values_to = "deaths") |> 
  mutate(age = as.numeric(str_remove(age, "x")))

# join these two tibbles and calculate rates

asmr <- d_long |> 
  left_join(dm_long) |> 
  mutate(mx = deaths/pop)

```

Do the decomposition of the difference between Kenya and Canada:

```{r}
asmr |> 
  # get data for kenya
  filter(region == "Kenya",year==2023)  |> 
  select(sex, age, pop, mx) |> 
  rename(pop_kenya = pop, mx_kenya = mx) |> 
  # get data for canada
  left_join(asmr |> 
              filter(region == "Canada",year==2023)  |> 
              select(sex, age, pop, mx) |> 
              rename(pop_can = pop, mx_can = mx) ) |> 
  # calculate population proportions
  mutate(prop_kenya = pop_kenya/sum(pop_kenya),
         prop_can = pop_can/sum(pop_can)) |> 
  mutate(rate_diff = mx_kenya - mx_can, # difference in mortality rates
         prop_diff = prop_kenya - prop_can) |> # diference in the pop proportions
  mutate(ave_rate = (mx_kenya+mx_can)/2, # average mortality
         ave_prop = (prop_kenya+prop_can)/2) |> # average pop proportion
  mutate(age_contr = prop_diff*ave_rate, # age contribution
         rate_contr = rate_diff*ave_prop) |> # mortality contribution
  summarize(age_total_contr = sum(age_contr), # summing up over age
            rate_total_contr = sum(rate_contr)) |> 
  mutate(total_diff = age_total_contr+rate_total_contr)
```

Check that the difference is actually the difference between the two CDRs

```{r}
asmr |> 
  filter(region == "Kenya"|region=="Canada",year==2023) |> 
  group_by(region) |> 
  summarize(cdr = sum(mx*pop)/sum(pop)) |> 
  summarise(diff = cdr[region=="Kenya"] - cdr[region=="Canada"])
```

## Exercise

Decompose the difference in CDRs between USA and Japan in the year 2023. Is the majority of the difference due to age structure or mortality?

```{r}
asmr |> 
  # get data for usa
  filter(region == "United States of America",year==2023)  |> 
  select(sex, age, pop, mx) |> 
  rename(pop_usa = pop, mx_usa = mx) |> 
  # get data for jpn
  left_join(asmr |> 
              filter(region == "Japan",year==2023)  |> 
              select(sex, age, pop, mx) |> 
              rename(pop_jpn = pop, mx_jpn = mx) ) |> 
  # calculate population proportions
  mutate(prop_usa = pop_usa/sum(pop_usa),
         prop_jpn = pop_jpn/sum(pop_jpn)) |> 
  mutate(rate_diff = mx_usa - mx_jpn, # difference in mortality rates
         prop_diff = prop_usa - prop_jpn) |> # diference in the pop proportions
  mutate(ave_rate = (mx_usa+mx_jpn)/2, # average mortality
         ave_prop = (prop_usa+prop_jpn)/2) |> # average pop proportion
  mutate(age_contr = prop_diff*ave_rate, # age contribution
         rate_contr = rate_diff*ave_prop) |> # mortality contribution
  summarize(age_total_contr = sum(age_contr), # summing up over age
            rate_total_contr = sum(rate_contr)) |> 
  mutate(total_diff = age_total_contr+rate_total_contr)
```
```{r}
asmr |> 
  filter(region == "United States of America"|region=="Japan",year==2023) |> 
  group_by(region) |> 
  summarize(cdr = sum(mx*pop)/sum(pop)) |> 
  summarise(diff = cdr[region=="United States of America"] - cdr[region=="Japan"])
```



# Mortality models

Read in mortality rates for Ontario. These data come from the [Canadian Human Mortality Database](http://www.bdlc.umontreal.ca/chmd/prov/ont/ont.htm). 

```{r}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
head(dm)
```

## Gompertz

Let's fit a Gompertz model to Male mortality rates in the year 1950 from age 40. What is the interpretation of the coefficient estimates?

```{r}
# clean up a bit
dm <- dm |> 
  clean_names() |> 
  mutate(age = as.numeric(age))

df_1950_40 <- dm |> 
  filter(year==1950, age>39) |> 
  select(age, male) 

df_1950_40 |> 
  ggplot(aes(age, male)) + 
  geom_point()+
  scale_y_log10()

# remove above 100
df_1950_40 <- df_1950_40 |> 
  filter(age<100)

mod <- lm(log(male)~age, data = df_1950_40)

# mortality at age 40 (estimated)
exp(coef(mod)[1])

summary(mod)
```

### Exercise

Now fit a Gompertz model to male mortality rates form age 40 in every year. Plot the estimated alpha and beta coefficients in a scatter plot, color the points by year. Comment on what you observe. 

```{r}
dm_to_fit <- dm |> 
  filter(age>39, age<100) |> 
  select(year, age, male)

coefs <- tibble()

years <- unique(dm_to_fit$year)

for(i in 1:length(years)){
  this_df <- dm_to_fit |> 
    ####SOMETHING
  
  this_mod <- lm(log(male)~age, data = this_df)
  
  these_coefs <- tibble(alpha = exp(coef(mod)[1]), beta = coef(mod)[2])
  
  coefs <- bind_rows(coefs, these_coefs)
  
}


```


## Lee-Carter


Let's get the Lee-carter model parameters for Ontario. First, get the matrix of age-specific rates:

```{r}
m_tx <- dm |> 
  filter(age < 101) |> 
  select(year, age, male) |> 
  pivot_wider(names_from = "age", values_from = "male") |> 
  select(-year) |> 
  as.matrix()

ages <- 0:100
years <- unique(dm$year)
```

log and demean those rates:

```{r}
logm_tx <- log(m_tx)
logm_tx[is.infinite(logm_tx)] <- min(logm_tx[!is.infinite(logm_tx)])
ax <- apply(logm_tx, 2, mean)

plot(ax)
```

Do the SVD

```{r}
# demeaning 
swept_logm_tx <- sweep(logm_tx, 2, ax)

svd_mx <- svd(swept_logm_tx)

names(svd_mx)


bx <- svd_mx$v[, 1]/sum(svd_mx$v[, 1])
kt <- svd_mx$d[1] * svd_mx$u[, 1] * sum(svd_mx$v[, 1])

```

Plots!

```{r}
# plot ax

lc_age_df <- tibble(age = ages, ax = ax, bx = bx)
lc_time_df <- tibble(year = years, kt = kt)

ggplot(lc_age_df, aes(age, ax)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("ax values for Ontario")

ggplot(lc_age_df, aes(age, bx)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("bx values for Ontario")

ggplot(lc_time_df, aes(year, kt)) + 
  geom_line(lwd = 1.1) + 
  ggtitle("kt values for Ontario")

```

let's look at the fit for a couple of years

```{r}
data_and_res <- dm |> 
  filter(age < 101) |> 
  mutate(log_mx = log(male)) |> 
  left_join(lc_age_df) |> 
  left_join(lc_time_df) |> 
  mutate(lc_fit = ax + bx*kt)

data_and_res |> 
  filter(year %in% c(1921, 1951, 1981, 2011, 2021)) |> 
  ggplot(aes(age, log_mx)) + geom_point() + 
  facet_wrap(~year) + 
  geom_line(aes(age, lc_fit), color = "red") + 
  ggtitle("Data and fits for Ontario")

```

### Exercise

Repeat the lee-carter model fitting exercise but just use mortality rates from 1970. Does this change the estimated rates? Does it do a better or worse job, or does it depend on the year? 




