---
title: "Week 7: Kinship"
subtitle: "SOC6708 ADA"
author: "Monica Alexander"
format: pdf
execute: 
  warning: false
  message: false
---

In this lab we will be calculating surviving daughters and mothers using the GKP equations, and also exploring matrix-based models using `DemoKin`.

# Load in data

Note you will have to install the DemoKin package.
```{r}
library(tidyverse)
library(readxl)
library(DemoKin)
```

For the GKP we will be using fertility and mortality rates from WPP; we also need to load in the file that has summary indicators. You should have these files from previous weeks. 

```{r}
df <- read_csv("../data/WPP2024_Fertility_by_Age5.csv")
dm <- read_csv("../data/WPP2024_Life_Table_Abridged_Medium_1950-2023.csv")
di <- read_csv("../data/WPP2024_Demographic_Indicators_Medium.csv")
```

For the matrix models, we will use fertility and mortality rates by single year of age. I've already cleaned these up (the file is in the repo).

```{r}
ds <- read_csv("../data/wpp_pop_fx_px_female.csv")
```

# Goodman-Keyfitz-Pullum equations

## Surviving daughters

Let's look at the surviving daughters in Kenya over time. 

First, clean up the fertility and mortality data:
```{r}
fert <- df |> 
  filter(Location == "Kenya") |> 
  select(Time, AgeGrp, ASFR) |> 
  mutate(age = as.numeric(str_remove(AgeGrp, "-.*")))

mort <- dm |> 
  filter(Location == "Kenya", Sex == "Female") |> 
  select(Time, AgeGrp, lx) |> 
  filter(AgeGrp!="1-4")|> # don't need this
  mutate(age = as.numeric(str_remove(AgeGrp, "-.*")))
```

### Exercise

Do a couple of graphs to show how fertility and mortality have changed over time. 


```{r}
fert |> 
  filter(Time %in% seq(1950, 2020, by = 10)) |> 
  ggplot(aes(age, ASFR, color = factor(Time))) +
  geom_line()+
  scale_color_brewer(name = "Year")+
  labs("y = Age-specific fertility rate", title = "Fertility over time in Kenya")
```


Let's calculate expected number of surviving daughters over time for a 60 year old:

```{r}
a <- 60

surviving_daughters <- fert |> 
  mutate(shifted_age = a - age) |> 
  left_join(mort |> rename(shifted_age = age) |> 
              select(-AgeGrp) ) |> 
  mutate(prod = ASFR*5*0.4886/1000*lx/100000) |> 
  group_by(Time) |> 
  summarize(daughters = sum(prod)) |> 
  filter(Time < 2024)
```

Plot this:

```{r}
surviving_daughters |> 
  ggplot(aes(Time, daughters)) + 
  geom_line()+
  labs(y = "Surviving daughters", title = "Expected number of surviving daughters for a 60 year woman, Kenya")
```
### Exercise

Compare the trajectory above to the trajectory of TFR for Kenya. Explain the similarities / differences.

```{r}
fert |> 
  group_by(Time) |> 
  summarize(TFR = sum(ASFR*5/1000)) |> 
  filter(Time < 2024) |> 
  ggplot(aes(Time, TFR))+
  geom_line()
```



## Surviving mothers

To calculate the probability that a mother is still alive, we need to calculate $W_x$, which is the age distribution of mothers. We'll calculate this using stable population assumptions as in the lecture notes, which requires first calculating the stable growth rate $r$. I'm using the approximation from week 5, i.e.
$$
r \approx \frac{NRR}{\mu}
$$

where $\mu$ is the mean age at childbearing. 

```{r}
dr <- di |> 
  filter(Location=="Kenya") |> 
  select(Time, NRR, MAC) |> 
  mutate(r = NRR/MAC)

Wx <- mort |> 
  left_join(fert) |> 
  mutate(ASFR = ifelse(is.na(ASFR), 0, ASFR)) |> 
  mutate(age = ifelse(is.na(age), 100, age)) |> 
  mutate(prod = ASFR*5*0.4886/1000*lx/100000) |> 
  left_join(dr) |> 
  mutate(dist = exp(-1*r*age)*prod) |> 
  group_by(Time) |> 
  mutate(sum = sum(dist)) |> 
  mutate(prop = dist/sum) # renormalize

```

Plot a couple of years to make sure it broadly makes sense

```{r}
Wx |> 
  filter(Time == 2023|Time == 1950) |> 
  ggplot(aes(age, prop, fill = factor(Time))) + 
  geom_bar(stat = "identity", position = 'dodge')+
  scale_fill_brewer(palette = "Set1", name = "Year")+
  labs(y = "Proportion of mothers", title = "Age distribution of mothers, Kenya")

```
Now we have this age distribution, we can calculate the probability that a mother is alive. For this one, let's use ego age = 45.

```{r}

a <- 45

age_shifted = seq(0, 100, by = 5) + a

# shift mortality rates
mort_shift <- mort |> 
  mutate(age = ifelse(is.na(age), 100, age)) |> 
  filter(age %in% age_shifted) |> 
  rename(lx_shifted = lx, age_shifted = age) |> 
  select(-AgeGrp)

surviving_mothers <- mort |> 
  mutate(age_shifted = age + a) |> 
  left_join(mort_shift) |> 
  mutate(lx_shifted = ifelse(is.na(lx_shifted), 0, lx_shifted)) |> 
  left_join(fert) |> 
  mutate(ASFR = ifelse(is.na(ASFR), 0, ASFR)) |> 
  left_join(Wx |> 
              select(Time, AgeGrp, age, prop)) |> 
  mutate(ratio = lx_shifted/lx) |> 
  mutate(prod = prop*ratio) |> 
  drop_na(age) |> 
  group_by(Time) |> 
  summarize(M1 = sum(prod)) 
```

```{r}
surviving_mothers|> 
  ggplot(aes(Time, M1)) + 
  geom_line()+
  labs(y = "Probability", title = "Probability that mother is still alive for Kenyan woman aged 45")
```
### Question for you

How would you convert this probability into the number of 45 year old women in Kenya who have lost their mother (say, in 2023)?

# DemoKin

Now we will use the `DemoKin` package to calculate surviving kin of different types (female only) in Canada. This code replicates the plots in the lecture notes. 

Let's get the fertility and mortality rates in the right format:

```{r}
asfr_2023 <- ds |> 
  filter(region=="Canada") |> 
  filter(year==2023) |> 
  select(age, year, fx) |> 
  mutate(fx = fx/1000) |> 
  pivot_wider(names_from = year, values_from = fx) |>
  select(-age) |>
  as.matrix()

surv_2023 <- ds |> 
  filter(region=="Canada") |> 
  filter(year==2023) |> 
  select(age, year, px) |> 
  pivot_wider(names_from = year, values_from = px) |>
  select(-age) |>
  as.matrix()
```

The main function in the DemoKin package is then `kin`, which takes an inputs fertility and survival matrices. 

```{r}
kin_2023 <- kin(p = surv_2023, f = asfr_2023, time_invariant = TRUE)
```

This object has two elements, a full description and a summary. 

```{r}
kin_2023$kin_full
unique(kin_2023$kin_full[,"kin"])
kin_2023$kin_summary
```

Note that this kin column refers to different types of kin, to make this more readable we can use the `rename_kin` function:

```{r}
kin_2023$kin_summary |> 
  rename_kin()
```
For example, let's look at the number of surviving daughters to 60 year olds:

```{r}
kin_2023$kin_summary |> 
  rename_kin() |> 
  filter(kin_label == "Daughters", age_focal==60)
```
We can plot a tree summary for a 45 year old:

```{r}
kin_2023$kin_summary |> 
  filter(age_focal == 45) |> 
  select(kin, count = count_living) |> 
  plot_diagram(rounding = 2)
```

We can specify the types of kin we are interested in. 

```{r}
kin_2023 <- 
  kin(
    p = surv_2023
    , f = asfr_2023
    , output_kin = c("c", "d", "gd", "ggd", "gm", "m", "n", "a", "s")
    , time_invariant = TRUE
  )
```

Then plot the number surviving over age

```{r}
kin_2023$kin_summary |>
  rename_kin() |>
  ggplot() +
  geom_line(aes(age_focal, count_living))  +
  theme_bw() +
  labs(x = "Age of focal", y= "Number of living female relatives") +
  facet_wrap(~kin_label)
```

Combine these counts:

```{r}
counts <- 
  kin_2023$kin_summary |>
  group_by(age_focal) |> 
  summarise(count_living = sum(count_living)) |> 
  ungroup()

kin_2023$kin_summary |>
  select(age_focal, kin, count_living) |> 
  rename_kin() |> 
  ggplot(aes(x = age_focal, y = count_living)) +
  geom_area(aes(fill = kin_label), colour = "black") +
  geom_line(data = counts, linewidth = 2) +
  labs(x = "Age of focal",
       y = "Number of living female relatives",
       fill = "Kin") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Finally, we can plot the age distribution of relatives for Canadian women aged 65.

```{r}
kin_2023$kin_full |>
  rename_kin() |>
  filter(age_focal == 25) |>
  ggplot(aes(age_kin, living)) +
  geom_line() +
  geom_vline(xintercept = 25, colour = "red") +
  labs(x = "Age of kin",
       y = "Expected number of living relatives") +
  theme_bw() +
  facet_wrap(~kin_label)
```

Can repeat the same exercise using time varying rates:

```{r}
asfr <- ds |> 
  filter(region=="Canada") |> 
  select(age, year, fx) |> 
  mutate(fx = fx/1000) |> 
  pivot_wider(names_from = year, values_from = fx) |>
  select(-age) |>
  as.matrix()

surv <- ds |> 
  filter(region=="Canada") |> 
  select(age, year, px) |> 
  pivot_wider(names_from = year, values_from = px) |>
  select(-age) |>
  as.matrix()

kin_tv <- kin(p = surv, f = asfr, time_invariant = FALSE, 
              output_kin = c("c", "d", "gd", "ggd", "gm", "m", "n", "a", "s"))

kin_tv$kin_full |> 
  filter(year==2023) |> 
  rename_kin() |>
  filter(age_focal == 65) |>
  ggplot(aes(age_kin, living)) +
  geom_line() +
  geom_vline(xintercept = 65, colour = "red") +
  labs(x = "Age of kin",
       y = "Expected number of living relatives") +
  theme_bw() +
  facet_wrap(~kin_label)
```

### Exercise

Pick another country to get the kinship structure for women. Compare to Canada and briefly discuss similarities and differences. 