---
title: "Week 9: Intro to Bayes"
date: today
date-format: "DD/MM/YY"
format: pdf
execute: 
  warning: false
  message: false
---

# Happiness

## Question 1

Consider the happiness example from the lecture, with 118 out of 129 women indicating they are happy. We are interested in estimating $\theta$, which is the (true) proportion of women who are happy. Calculate the MLE estimate $\hat{\theta}$ and 95% confidence interval. 

```{r}
y <- 118
n <- 129

# mean
y/n
```

```{r}
p <- y/n
se <- sqrt(p*(1-p)/n)

p - 1.96*se
p + 1.96*se
```


## Question 2

Assume a Beta(1,1) prior on $\theta$. Calculate the posterior mean for $\hat{\theta}$ and 95% credible interval. 

```{r}
# prior Beta (1,1) == Uniform(0,1)
# posterior Beta(y+1, n-y+1)
a <- 1
b <- 1

p_bayes <- (y+a)/(n+a+b)
p_bayes

qbeta(0.025, y+a, n-y+b)
qbeta(0.975, y+a, n-y+b)
```


## Question 3

Now assume a Beta(10,10) prior on $\theta$. What is the interpretation of this prior? Are we assuming we know more, less or the same amount of information as the prior used in Question 2?

```{r}
# prior Beta (1,1) == Uniform(0,1)
# posterior Beta(y+1, n-y+1)
a <- 10
b <- 10

p_bayes <- (y+a)/(n+a+b)
p_bayes

qbeta(0.025, y+a, n-y+b)
qbeta(0.975, y+a, n-y+b)
```




```{r}
# prior Beta (1,1) == Uniform(0,1)
# posterior Beta(y+1, n-y+1)
a <- 10
b <- 100

p_bayes <- (y+a)/(n+a+b)
p_bayes

qbeta(0.025, y+a, n-y+b)
qbeta(0.975, y+a, n-y+b)
```

```{r}
library(tidyverse)
dp <- read_csv("../data/wpp_pop_clean.csv")
df <- read_csv("../data/WPP2024_Demographic_Indicators_Medium.csv")

dpf <- dp |> filter(region=="Canada", sex == "Female") |> 
  pivot_longer(x0:x100) |> 
  mutate(age = as.numeric(str_remove(name, "x")))

dpm <- dp |> filter(region=="Canada", sex == "Male") |> 
  pivot_longer(x0:x100) |> 
  mutate(age = as.numeric(str_remove(name, "x")))

breaks <- c(seq(0,100, by = 5), Inf)
labels <- seq(0,100, by = 5)

# Create the age groups
dpf$age_group <- cut(dpf$age, 
                  breaks = breaks, 
                  labels = labels, 
                  right = FALSE) 

dpf5 <- dpf |> 
  group_by(year, age_group) |> 
  summarize(pop = sum(value))

dpm$age_group <- cut(dpm$age, 
                  breaks = breaks, 
                  labels = labels, 
                  right = FALSE) 

dpm5 <- dpm |> 
  group_by(year, age_group) |> 
  summarize(pop = sum(value))

dp5_23 <- dpf5 |> filter(year==2023)

C <- dp5_23$pop[dp5_23$age_group==0] + dpm5$pop[dpm5$age_group==0&dpm5$year==2023]

W <- dp5_23$pop[dp5_23$age_group==15]+dp5_23$pop[dp5_23$age_group==20]+dp5_23$pop[dp5_23$age_group==25]+dp5_23$pop[dp5_23$age_group==30]+dp5_23$pop[dp5_23$age_group==35]+dp5_23$pop[dp5_23$age_group==40]+dp5_23$pop[dp5_23$age_group==45]

p <- 1/7
s <- 1

iTFR <- 1/s*1/p*C/W
iTFR

x_multiplier <- (10.65-12.55*(dp5_23$pop[dp5_23$age_group==25]+dp5_23$pop[dp5_23$age_group==30])/W)
xTFR <- x_multiplier*C/W
xTFR

# the actual TFR is
df |> filter(Location=="Canada") |> select(Time, TFR)
```

